The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Innate nature or tidal nurture?

### 2025-10-21 thesis/extra_numerical_convergence.tex
thesis updates
thesis text to kim

Due to the finite resolution of N-body simulations, collisional (close)
encounters between particles are inevitable. Such collisional encounters
are both unphysical (with possible arbitrary acceleration, violating the
``collisionless'' assumption) and computationally expensive (requiring
many small time-steps). As a remedy, many N-body codes use a
``softened'' gravitational force law, where the force weakens when
closer to a particle than a ``softening length''. The choice of
softening length ideally balances resolution and computational speed.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) particles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar(s).
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths than the fiducial. \textbf{Bottom:} Models with more precise
timestep accuracy and gravitational force accuracy. \textbf{Summary:}
Except for the model with a larger softening length, all simulations
agree within uncertainties.}\label{fig:methods_convergence}
\end{figure}

>>>>

Due to the finite resolution of N-body simulations, collisional (close)
encounters between particles are inevitable. Such collisional encounters
are both unphysical (with possible arbitrary acceleration, violating the
``collisionless'' assumption) and computationally expensive (requiring
many small time-steps). As a remedy, many N-body codes use a
\emph{softened} gravitational force law, where the force weakens when
closer to a particle than a \emph{softening length.} The choice of
softening length ideally balances resolution and computational speed.

@@

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/orbit_converg_methods.png}
\caption[Numerical methods convergence]{Similar to
Fig.~\ref{fig:numerical_convergence}, except a comparison of the final
velocity profiles using different simulation methods for Sculptor's
\smallperi{} model. The benchmark model is our fiducial \(10^7\)
particle run, and all other models use \(10^5\) particles, with their
``converged radius'' marked by the black arrow and softening by the
vertical bar(s). \textbf{Left:} Models with \(\sqrt{10}\) larger and
smaller softening lengths than the fiducial. \textbf{Right:} Models with
more precise timestep accuracy and gravitational force
accuracy.}\label{fig:methods_convergence}
\end{figure}

### 2025-10-21 thesis/extra_results.tex
thesis text to kim

Changing the initial concentration (or \(\vmax\) and \(\rmax\))
primarily affects total dark matter evolution.
Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
\smallperi{} orbit with different initial halo parameters. The heavier
halo has \(\vmax=43\) and \(\rmax=7\), whereas the lighter halo has
\(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
more mass, the halo evolves to similar final structural parameters.

@@

Radial velocity anisotropy may cause halos to disrupt faster
\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
velocity anisotropy, we initialize a model with a velocity anisotropy
with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
to \(\beta=1\) and infinity, with scale length \(4r_s\).

@@

We create an oblate initial dark matter halo using \agama{}.
Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
isolation) isodensity contours of our model. The initial snapshot is
sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor of 0.5 in the \(z\)-axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo projected density snapshots]{The initial and final
(after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, can approximately reproduce the
observed position of Scl (see Fig.~\ref{fig:scl_mw_impact_orbit}). The
model has a Galactocentric initial position of
\([67.83, -352.2, 110.3]\,\kpc\) and velocity of
\([-3.68, 30.79, -22.77]\,\kpc\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_orbits.pdf}}
\caption[Sculptor MW impact orbit]{The orbit of Sculptor for the
MW-impact model. The point (dotted) and n-body (solid) diverge by
\(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
\end{figure}

@@

Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream} show
the resulting distributions of velocities and distance along the stream
orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
such a gradient would require tracing stars across several degrees on
the sky.

@@

While we have compared the evolution of orbits in the MW potential for
simplicity, because the tidal evolution of Scl is very weak in the LMC
model, the differences between models become even more slight. And while
UMi has a lower pericentre than Scl, the differences between halo
structure should apply similarly to UMi.

>>>>

Changing the initial concentration (or \(\vmax\) and \(\rmax\))
primarily affects total dark matter evolution.
Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
\smallperi{} orbit with different initial halo parameters. The heavier
halo has \(\vmax=43\,\kms\) and \(\rmax=7\,\kpc\), whereas the lighter
halo has \(\vmax=25\,\kms\) and \(\rmax=2.5\,\kpc\). While the less
concentrated halo loses more mass, the halo evolves to similar final
structural parameters.

@@

Velocity anisotropy quantifies the number of radial versus circular
orbits. Specifically, the anisotropy parameter,
\(\beta = 1 - \frac{\sigma_{\phi}^2 + \sigma_\theta^2}{2\sigma_{r}^2}\)
where \(\sigma^2\) is the velocity dispersion in the spherical \(r\),
\(\phi\), or \(\theta\) direction. Radial velocity anisotropy may cause
halos to disrupt faster \citep[e.g.][]{chiang+bosch+schive2024}. To test
the effects of moderate velocity anisotropy, we initialize a model with
a velocity anisotropy with an Osipkov-Merritt profile rising from
\(\beta=0.2\) at the centre to \(\beta=1\) and infinity, with scale
length \(4r_s\).

@@

We create an oblate initial dark matter halo using \agama{}.
Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after
\(5\,\Gyr\) in isolation) isodensity contours of our model. The initial
snapshot is sampled from the distribution function \begin{equation}{
\begin{split}
f({\bf J}) = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h({\bf J})}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g({\bf J})}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g({\bf J})}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g({\bf J}) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h({\bf J}) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor of 0.5 in the \(z\)-axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo projected density snapshots]{The initial and final
(after \(9\,\Gyr\) of tidal evolution) projected density profiles for
the oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\)
and \(z'\) the major and minor axes. The contours are drawn assuming
normal smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, can approximately reproduce the
observed position of Scl (see Fig.~\ref{fig:scl_mw_impact_orbit}). The
model has a Galactocentric initial position of
\([67.83, -352.2, 110.3]\,\kpc\) and velocity of
\([-3.68, 30.79, -22.77]\,\kms\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_orbits.pdf}}
\caption[Sculptor MW impact orbit]{Similar to
Fig.~\ref{fig:scl_lmc_orbits_effect}, except for the orbit of Sculptor
for the MW-impact model. The point (dotted) and N-body (solid) diverge
by \(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
\end{figure}

@@

Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream} show
the resulting distributions of velocities and distance along the stream
orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
detectible gradients in radial velocities (\(\sim 10\,\kms\) over 5
degrees) and proper motions (mostly for UMi, of \(\sim 0.1\,\masyr\)
over 5 degrees). However, detecting such a gradient would require
tracing stars across several degrees on the sky.

@@

While we compared the evolution of orbits of Scl in the MW-only
potential for simplicity, we expect the differences between models in
the MW and LMC potential to be more slight due to the weaker tidal
evolution. And while UMi has a lower pericentre than Scl, the
differences between halo structure should apply similarly to UMi. We
conclude that our assumptions regarding orbits, halo concentration,
cuspyness, anisotropy, and ellipticity likely do not dramatically change
our conclusions.

### 2025-10-21 thesis/extra_rv_models.tex
thesis updates
thesis text to kim

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[Velocity dispersion fits]{\textbf{Left:} Velocity histogram of
Scl and UMi in terms of projected-correction GSR LOS velocity. Black
points with error bars are from the sample of LOS-velocity members, and
green lines represent MCMC fits to the velocity dispersion.
\textbf{Right:} The observed velocity dispersion gradient (black points)
in 10 equal number bins and the derived slopes for MCMC samples (green
lines). Scl shows moderate evidence for an increasing velocity
dispersion with radius, but Scl's velocity dispersion appears to flatten
outside of \(R_h\). UMi's evidence for a radial gradient is
weaker.}\label{fig:rv_hists}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
\caption[A possible velocity gradient in Sculptor]{\textbf{Top} members
of Sculptor plotted in the tangent plane, coloured by corrected velocity
difference from mean \(v_z - \bar v_z\). The black ellipse marks the
half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
arrows mark the proper motion (PM, GSR frame) and derived velocity
gradient (rot) vectors (to scale). \textbf{Bottom}: The corrected LOS
velocity along the most likely rotational axis. RV members are black
points, the systematic \(v_z\) is the horizontal grey line, blue lines
represent the (projected) gradient from MCMC samples, and the orange
line is a rolling median (with a window size of
50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

>>>>

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[Velocity dispersion fits]{\textbf{Left:} Velocity histogram of
Scl and UMi in terms of projected-correction LOS velocity. Black points
with error bars are a histogram of LOS-velocity members, and green lines
represent MCMC fits to the velocity dispersion. \textbf{Right:} The
observed velocity dispersion gradient (black points) in 10 equal number
bins and the derived slopes for MCMC samples (green lines). Scl shows
moderate evidence for an increasing velocity dispersion with radius, but
Scl's velocity dispersion appears to flatten outside of \(R_h\). UMi's
evidence for a radial gradient is weaker.}\label{fig:rv_hists}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.pdf}}
\caption[A possible velocity gradient in Sculptor]{\textbf{Top} members
of Sculptor plotted in the tangent plane, coloured by corrected velocity
difference from mean \(v_{\rm gsr}' - \bar v_{\rm gsr}'\). The black
ellipse marks the half-light radius in Fig.~\ref{fig:scl_selection}. The
black and green arrows mark the proper motion (PM, GSR frame) and
derived velocity gradient (rot) vectors (to scale). \textbf{Bottom}: The
corrected LOS velocity along the most likely rotational axis. RV members
are black points, the systematic \(v_{\rm gsr}''\) is the horizontal
grey line, blue lines represent the (projected) gradient from MCMC
samples, and the orange line is a rolling median (with a window size of
50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_rv_scatter_gradient.pdf}}
\caption[velocity]{Similar to
Fig.~\ref{fig:scl_velocity_gradient_scatter} except for Ursa Minor. Ursa
Minor does not have clear evidence of a velocity
gradient.}\label{fig:umi_velocity_gradient_scatter}
\end{figure}
